---
title: "1. RPrograms"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. PRINT HELLO WORLD
```{r}
# We can use the print() function
print("Hello World!")
#[1] "Hello World!"

# Quotes can be suppressed in the output
print("Hello World!", quote = FALSE)
# [1] Hello World!

# If there are more than 1 item, we can concatenate using paste()
print(paste("How","are","you?"))
# [1] "How are you?"

print(cat("How","are","you?"))

```


#2 ADDING VECTORS
```{r}

x <- c(3,6,8)
x
# [1] 3 6 8

y <- c(2,9,0)
y
# [1] 2 9 0

x + y
# [1]  5 15  8

x + 1    # 1 is recycled to (1,1,1)
# [1] 4 7 9

x + c(1,4)    # (1,4) is recycled to (1,4,1) but warning issued
# [1]  4 10  9
# Warning message:
# In x + c(1, 4) :
# longer object length is not a multiple of shorter object length
```


# 3 Find Sum, Mean and Product of Vector in R Programming

Example: Vector Elements Arithmetic
```{r}

sum(2,7,5)
[1] 14

x <- c(2, NA,  3,  1,  4)
x
# [1]  2 NA  3  1  4

sum(x)    # if any element is NA or NaN, result is NA or NaN
# [1] NA

sum(x, na.rm=TRUE)    # this way we can ignore NA and NaN values
# [1] 10

mean(x, na.rm=TRUE)
# [1] 2.5

prod(x, na.rm=TRUE)
# [1] 24

```


# 4. R Program to Take Input From User
```{r}
my.name <- readline(prompt="Enter name: ")
my.age <- readline(prompt="Enter age: ")

# convert character into integer
my.age <- as.integer(my.age)

print(paste("Hi,", my.name, "next year you will be", my.age+1, "years old."))

```



# 5.  R Program to Generate Random Number from Standard Distributions

Example: Uniform Distribution

```{r}
runif(1)    # generates 1 random number
# [1] 0.3984754

runif(3)    # generates 3 random number
# [1] 0.8090284 0.1797232 0.6803607

runif(3, min=5, max=10)    # define the range between 5 and 10
# [1] 7.099781 8.355461 5.173133
```


From Normal Distribution
```{r}
rnorm(1)    # generates 1 random number
# [1] 1.072712

rnorm(3)    # generates 3 random number
# [1] -1.1383656  0.2016713 -0.4602043

rnorm(3, mean=10, sd=2)    # provide our own mean and standard deviation
# [1]  9.856933  9.024286 10.822507
```


# 6 R Program to Sample from a Population

```{r}
x <- c(1,3,5,7,9,11,13,15,17) ; x
#[1]  1  3  5  7  9 11 13 15 17

# sample 2 items from x
sample(x, 2)
#[1] 13  9


# sample with replacement
sample(x, replace = TRUE)
#[1] 15 17 13  9  5 15 11 15  1

# if we simply pass in a positive number n, it will sample
# from 1:n without replacement
sample(10)
# [1]  2  4  7  9  1  3 10  5  8  6


sample(c("H","T"),10, replace = TRUE)
# [1] "H" "H" "H" "T" "H" "T" "H" "H" "H" "T"

```



# 7 R Program to Find Minimum and Maximum

```{r}
x <- c(5,8,3,9,2,7,4,6,10); x
# [1]  5  8  3  9  2  7  4  6 10

# find the minimum
min(x)
# [1] 2

# find the maximum
max(x)
# [1] 10

# find the range
range(x)
# [1]  2 10

```

If we want to find where the minimum or maximum is located, i.e. the index instead of the actual value, then we can use which.min() and which.max() functions.

```{r}
x <- c(5,8,3,9,2,7,4,6,10); x
# [1]  5  8  3  9  2  7  4  6 10

# find index of the minimum
which.min(x)
# [1] 5

# find index of the minimum
which.max(x)
# [1] 9

# alternate way to find the minimum
x[which.min(x)]
# [1] 2
```



# 8 R Program to Sort a Vector

```{r}
x <- c(7,1,8,3,2,6,5,2,2,4); x
# [1] 7 1 8 3 2 6 5 2 2 4

# sort in ascending order
sort(x)
# [1] 1 2 2 2 3 4 5 6 7 8

# sort in descending order
sort(x, decreasing=TRUE)
# [1] 8 7 6 5 4 3 2 2 2 1

# vector x remains unaffected
x
# [1] 7 1 8 3 2 6 5 2 2 4
```



Sometimes we would want the index of the sorted vector instead of the values. In such case we can use the function order()

```{r}
order(x)
# [1]  2  5  8  9  4 10  7  6  1  3

order(x, decreasing=TRUE)
# [1]  3  1  6  7 10  4  5  8  9  2

x[order(x)]    # this will also sort x
# [1] 1 2 2 2 3 4 5 6 7 8

```

# 9 R Program to Find the Factorial of a Number

```{r}
# take input from the user
num = as.integer(readline(prompt="Enter a number: "))
factorial = 1

# check is the number is negative, positive or zero
if(num < 0) {
    print("Sorry, factorial does not exist for negative numbers")
} else if(num == 0) {
    print("The factorial of 0 is 1")
} else {
    for(i in 1:num) {
        factorial = factorial * i
    }
    print(paste("The factorial of", num ,"is",factorial))
}
```



# 10. R Multiplication Table

```{r}
# Program to find the multiplication
# table (from 1 to 10)
# of a number input by the user

# take input from the user
num = as.integer(readline(prompt = "Enter a number: "))

# use for loop to iterate 10 times
for(i in 1:10) {
    print(paste(num,'x', i, '=', num*i))
}
```



# 11. R Program to Check Prime Number
```{r}
# Program to check if
# the input number is
# prime or not

# take input from the user
num = as.integer(readline(prompt="Enter a number: "))

flag = 0
# prime numbers are greater than 1
if(num > 1) {
    # check for factors
    flag = 1
    for(i in 2:(num-1)) {
        if ((num %% i) == 0) {
            flag = 0
            break
        }
    }
} 

if(num == 2)    flag = 1
if(flag == 1) {
    print(paste(num,"is a prime number"))
} else {
    print(paste(num,"is not a prime number"))
}
```

# 12 R Program to check Armstrong Number

An Armstrong number, also known as narcissistic number, is a number that is equal to the sum of the cubes of its own digits.

For example, 370 is an Armstrong number since 370 = 3*3*3 + 7*7*7 + 0*0*0.

```{r}
# take input from the user
num = as.integer(readline(prompt="Enter a number: "))

# initialize sum
sum = 0

# find the sum of the cube of each digit
temp = num
while(temp > 0) {
    digit = temp %% 10
    sum = sum + (digit ^ 3)
    temp = floor(temp / 10)
}

# display the result
if(num == sum) {
    print(paste(num, "is an Armstrong number"))
} else {
    print(paste(num, "is not an Armstrong number"))
}
```

# 13 R Program to Print the Fibonacci Sequence

A Fibonacci sequence is the integer sequence of

0, 1, 1, 2, 3, 5, 8....

```{r}
# take input from the user
nterms = as.integer(readline(prompt="How many terms? "))

# first two terms
n1 = 0
n2 = 1
count = 2

# check if the number of terms is valid
if(nterms <= 0) {
    print("Plese enter a positive integer")
} else {
    if(nterms == 1) {
        print("Fibonacci sequence:")
        print(n1)
    } else {
        print("Fibonacci sequence:")
        print(n1)
        print(n2)
        while(count < nterms) {
            nth = n1 + n2
            print(nth)
            # update values
            n1 = n2
            n2 = nth
            count = count + 1
        }
    }
}
```


# 14 R Program to Check for Leap Year

```{r}
# Program to check if
# the input year is
# a leap year or not

year = as.integer(readline(prompt="Enter a year: "))
if((year %% 4) == 0) {
    if((year %% 100) == 0) {
        if((year %% 400) == 0) {
            print(paste(year,"is a leap year"))
        } else {
            print(paste(year,"is not a leap year"))
        }
    } else {
        print(paste(year,"is a leap year"))
    }
} else {
    print(paste(year,"is not a leap year"))
}
```



# 15 Check if a Number is Odd or Even in R Programming

```{r}
# Program to check if
# the input number is odd or even.
# A number is even if division
# by 2 give a remainder of 0.
# If remainder is 1, it is odd.

num = as.integer(readline(prompt="Enter a number: "))
if((num %% 2) == 0) {
    print(paste(num,"is Even"))
} else {
    print(paste(num,"is Odd"))
}
```

# 16 R Program to Check if a Number is Positive, Negative or Zero
```{r}
# In this program, we input a number
# check if the number is positive or
# negative or zero and display
# an appropriate message

num = as.double(readline(prompt="Enter a number: "))
if(num > 0) {
    print("Positive number")
} else {
    if(num == 0) {
        print("Zero")
    } else {
        print("Negative number")
    }
}
```

# 17 R Program to Find the Sum of Natural Numbers

```{r}
# take input from the user
num = as.integer(readline(prompt = "Enter a number: "))

if(num < 0) {
    print("Enter a positive number")
} else {
    sum = 0

    # use while loop to iterate until zero
    while(num > 0) {
        sum = sum + num
        num = num - 1
    }

    print(paste("The sum is", sum))
}
```

From mathematics, we know that sum of natural numbers is given by

n*(n+1)/2

```{r}
# take input from the user
num = as.integer(readline(prompt = "Enter a number: "))

if(num < 0) {
    print("Enter a positive number")
} else {
    sum = (num * (num + 1)) / 2;

    print(paste("The sum is", sum))
}
```


# 18. Convert Decimal into Binary using Recursion in R

```{r}
# Program to convert decimal
# number into binary number
# using recursive function

convert_to_binary <- function(n) {

    if(n > 1) {
        convert_to_binary(as.integer(n/2))
    }
    cat(n %% 2)
}
```



# 19 R program to Find the Factorial of a Number Using Recursion
```{r}
recur_factorial <- function(n) {
    if(n <= 1) {
        return(1)
    } else { 
        return(n * recur_factorial(n-1))
    }
}
```


# 20. R Program to Find the Factors of a Number
```{r}
print_factors <- function(x) {
    print(paste("The factors of",x,"are:"))
    for(i in 1:x) {
        if((x %% i) == 0) {
            print(i)
        }
    }
}
```


# 21 Fibonacci Sequence Using Recursion in R

```{r}
# Program to diplay the Fibonacci
# sequence up to n-th term using
# recursive functions

recurse_fibonacci <- function(n) {
    if(n <= 1) {
        return(n)
    } else {
        return(recurse_fibonacci(n-1) + recurse_fibonacci(n-2))
    }
}

# take input from the user
nterms = as.integer(readline(prompt="How many terms? "))

# check if the number of terms is valid
if(nterms <= 0) {
    print("Plese enter a positive integer")
} else {
    print("Fibonacci sequence:")
    for(i in 0:(nterms-1)) {
        print(recurse_fibonacci(i))
    }
}
```

22. R Program to Find H.C.F. or G.C.D.

```{r}
# Program to find the
# H.C.F of two input number

# define a function
hcf <- function(x, y) {
    # choose the smaller number
    if(x > y) {
        smaller = y
    } else {
        smaller = x
    }
    for(i in 1:smaller) {
        if((x %% i == 0) && (y %% i == 0)) {
            hcf = i
        }
    }
    return(hcf)
}

# take input from the user
num1 = as.integer(readline(prompt = "Enter first number: "))
num2 = as.integer(readline(prompt = "Enter second number: "))

print(paste("The H.C.F. of", num1,"and", num2,"is", hcf(num1, num2)))
```



Euclidean algorithm to Find GCD
This algorithm is based on the fact that H.C.F. of two numbers divides their difference as well.

In this algorithm, we divide the greater by smaller and take the remainder. Now, divide the smaller by this remainder. Repeat until the remainder is 0.

For example, if we want to find the H.C.F. of 54 and 24, we divide 54 by 24. The remainder is 6.

Now, we divide 24 by 6 and the remainder is 0. Hence, 6 is the required H.C.F. We can do this in Python as follows.

Example 2: GCD Using Euclidean algorithm
```{r}
hcf <- function(x, y) {
    while(y) {
        temp = y
        y = x %% y
        x = temp
    }
    return(x)
}
```


# 23. R Program to Find L.C.M

Example: Compute LCM in R

```{r}
# Program to find the L.C.M. of two input number
lcm <- function(x, y) {
    # choose the greater number
    if(x > y) {
        greater = x
    } else {
        greater = y
    }

    while(TRUE) {
        if((greater %% x == 0) && (greater %% y == 0)) {
            lcm = greater
            break
        }
        greater = greater + 1
    }
    return(lcm)
}

# take input from the user
num1 = as.integer(readline(prompt = "Enter first number: "))
num2 = as.integer(readline(prompt = "Enter second number: "))

print(paste("The L.C.M. of", num1,"and", num2,"is", lcm(num1, num2)))

```


# 24. R Program to Make a Simple Calculator
```{r}
# Program make a simple calculator
# that can add, subtract, multiply
# and divide using functions

add <- function(x, y) {
    return(x + y)
}

subtract <- function(x, y) {
    return(x - y)
}

multiply <- function(x, y) {
    return(x * y)
}

divide <- function(x, y) {
    return(x / y)
}

# take input from the user
print("Select operation.")
print("1.Add")
print("2.Subtract")
print("3.Multiply")
print("4.Divide")

choice = as.integer(readline(prompt="Enter choice[1/2/3/4]: "))

num1 = as.integer(readline(prompt="Enter first number: "))
num2 = as.integer(readline(prompt="Enter second number: "))

operator <- switch(choice,"+","-","*","/")
result <- switch(choice, add(num1, num2), subtract(num1, num2), multiply(num1, num2), divide(num1, num2))

print(paste(num1, operator, num2, "=", result))
```



# 25. Sum of Natural Numbers Using Recursion

```{r}
# Program to find the sum of
# natural numbers upto n
# using recursive function

calculate_sum() <- function(n) {
    if(n <= 1) {
        return(n)
    } else {
        return(n + calculate_sum(n-1))
    }
}
```

