---
title: "2. DataWrangling , "Tidyr""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# tidyr

## What is tidyr?

tidyr is an R package that makes it easy to "tidy" your data frames.

It has a core set of "verbs" (functions) that handle most of our essential needs.

In this session, we'll look at the following:

- `unite()`
- `separate()`
- `gather()`
- `spread()`

## Getting tidyr
```{r, eval = F}

# Load it into your workspace
library(tidyr)
library(dplyr)

```

## Function Style

Just like functions in `dplyr`, the "verbs" of tidyr all share a similar format:

- their first argument is a data frame.
- subsequent arguments are specific to the function.
- they all return a data frame.

When referencing column names, we don't need quotation marks ""

# Time to tidy...

## Combine two variables into one with `unite()`

If you have two variables you'd like to combine into one, use `unite()`

```{r}
getwd()
# setwd("C:/Users/avina/Desktop/Interview/DataMining in R/0. dplyr-tutorial Project")
d <- "flight-times.csv" %>%
  read.csv()
d <- dplyr::tbl_df(d)
```

## 1. Combine two variables into one with `unite()`

Say we wanted to combine `carrier` and `fl_num` to get a single `flight` variable.

```{r}
dplyr::glimpse(d)
dat <- select(d, carrier, fl_num, origin, dest)
unite(dat, flight, carrier, fl_num)
unite(dat, flight, carrier, fl_num, sep = "") %>% head(2)
mod <- unite(dat, flight, carrier, fl_num, sep = "-")
head(mod, n = 3)

```

## 2. Separate a single column with `separate()`

Say all we had was `flight`, but wanted to get `carrier` and `fl_num`.

```{r}
separate(mod, flight, into = c("carrier", "flight_num"), sep = "-")
```



## 3. Gather multiple columns together with `gather()`

Convert multiple columns into two columns: a key and a value.

Useful for going from wide format to long format.

`gather(data, key.name, value.name, columns.to.gather)`


```{r}
d <- read.csv("energy.csv")
d <- dplyr::tbl_df(d)
dplyr::glimpse(d)

gather(d, exercise, energy, Running:Cycling)
```

## 4. Spread key-value columns with `spread()`

Convert two columns, representing a key and value, into multiple columns.

Useful for getting from long to wide format.

Say we started in long format:

```{r}
d <- gather(d, exercise, energy, Running:Cycling) 
head(d,n =2)
d %>% spread(exercise, energy)
```

