---
title: "2.DataWrangling_Dplyr"
output: html_document
---

## Brief Introduction to `dplyr`

The learning curve for R syntax is slow. One of the more difficult aspects that requires some getting used to is subsetting data tables. The `dplyr` package brings these tasks closer to English and we are therefore going to introduce two simple functions: one is used to subset and the other to select columns.

Take a look at the dataset we read in:
```{r}
getwd()
setwd("C:/Users/avina/Desktop/Edx/R & Statistics_MIT/Week 1")
filename <- "femaleMiceWeights.csv"
dat <- read.csv(filename)
head(dat) #In R Studio use View(dat)
```

There are two types of diets, which are denoted in the first column. If we want just the weights, we only need the second column. So if we want the weights for mice on the `chow` diet, we subset and filter like this:

```{r,message=FALSE}
library(dplyr) 
chow <- filter(dat, Diet=="chow") #keep only the ones with chow diet
head(chow)
```

And now we can select only the column with the values:

```{r}
chowVals <- select(chow,Bodyweight)
head(chowVals)
```

A nice feature of the `dplyr` package is that you can perform consecutive tasks by using what is called a "pipe". In `dplyr` we use `%>%` to denote a pipe. This symbol tells the program to first do one thing and then do something else to the result of the first. Hence, we can perform several data manipulations in one line. For example:

```{r}
chowVals <- filter(dat, Diet=="chow") %>% select(Bodyweight)
```

In the second task, we no longer have to specify the object we are editing since it is whatever comes from the previous call. 

Also, note that if `dplyr` receives a `data.frame` it will return a `data.frame`. 
```{r}
class(dat)
class(chowVals)
```

For pedagogical reasons, we will often want the final result to be a simple `numeric` vector. To obtain such a vector with `dplyr`, we can apply the `unlist` function which turns `lists`, such as `data.frames`, into `numeric` vectors:

```{r}
chowVals <- filter(dat, Diet=="chow") %>% select(Bodyweight) %>% unlist 
class( chowVals )
chowVals
```


To do this in R without `dplyr` the code is the following:

```{r}
chowVals <- dat[ dat$Diet=="chow", colnames(dat)=="Bodyweight"]
chowVals
```





---
title: "Introduction to dplyr for Faster Data Manipulation in R"
output: html_document
---

Note: There is a 40-minute [video tutorial](https://www.youtube.com/watch?v=jWjqLW-u3hc) on YouTube that walks through this document in detail.

## Why do I use dplyr?

* Great for data exploration and transformation
* Intuitive to write and easy to read, especially when using the "chaining" syntax (covered below)
* Fast on data frames


## dplyr functionality

* Five basic verbs: `filter`, `select`, `arrange`, `mutate`, `summarise` (plus `group_by`)
* Can work with data stored in databases and [data tables](http://datatable.r-forge.r-project.org/)
* Joins: inner join, left join, semi-join, anti-join (not covered below)
* Window functions for calculating ranking, offsets, and more
* [Better than plyr](http://blog.rstudio.org/2014/01/17/introducing-dplyr/) if you're only working with data frames (though it doesn't yet duplicate all of the plyr functionality)
* Examples below are based upon the [latest release](https://github.com/hadley/dplyr/releases), version 0.2 (released May 2014)




## Loading dplyr and an example dataset

* dplyr will mask a few base functions
* If you also use plyr, load plyr first
* hflights is flights departing from two Houston airports in 2011

```{r}
# load packages
suppressMessages(library(dplyr))
library(hflights)

# explore data
data(hflights)
head(hflights)
```

* `tbl_df` creates a "local data frame"
* Local data frame is simply a wrapper for a data frame that prints nicely

```{r}
# convert to local data frame
flights <- tbl_df(hflights)

# printing only shows 10 rows and as many columns as can fit on your screen
flights
```

```{r results='hide'}
# you can specify that you want to see more rows
print(flights, n=20)
head(flights)

# convert to a normal data frame to see all of the columns
data.frame(head(flights))


# To look at the structure of the dataframe (Using R)
str(flights)

# Using dplyr
glimpse(flights)

# Creating a dplyr dataframe
dplyr::data_frame(a = 1:3, b = 4:6)
?data_frame


str(tibble("letters"))

```


```{r}
# to view the dataframe
utils::View(flights)
View(flights)

```





## Subset Observations (Rows) - Selecting Rows

## filter: Keep rows matching criteria

* Base R approach to filtering forces you to repeat the data frame's name
* dplyr approach is simpler to write and read
* Command structure (for all dplyr verbs):
    * first argument is a data frame
    * return value is a data frame
    * nothing is modified in place
* Note: dplyr generally does not preserve row names

```{r results='hide'}
# base R approach to view all flights on January 1
flights[flights$Month==1 & flights$DayofMonth==1, ]
```

```{r}
# dplyr approach
# note: you can use comma or ampersand to represent AND condition
filter(flights, Month==1, DayofMonth==1)

# use pipe for OR condition
filter(flights, UniqueCarrier=="AA" | UniqueCarrier=="UA")

dplyr::filter(flights, Year < 2012 & DepTime < 0200 & Month == 2)
```

```{r results='hide'}
# you can also use %in% operator
filter(flights, UniqueCarrier %in% c("AA", "UA"))
```



```{r}
# distinct function from dplyr

flights
dplyr::distinct(flights, Year)
dplyr::distinct(flights, UniqueCarrier, FlightNum) %>% dplyr::arrange(FlightNum)

```



```{r}
# slice function from dplyr

utils::head(flights)
dplyr::slice(flights, 3:5)
dplyr::slice(flights, 2)

```



```{r}

# Randomly select n rows.
# randomly sample a fixed number of rows, without replacement
flights %>% sample_n(5)


# Randomly select fraction of rows.
# randomly sample a fraction of rows, with replacement
flights %>% sample_frac(0.25, replace=TRUE)



```







## Subset Variables (Columns) - Selecting columns

## select: Pick columns by name

```{r}
head(flights, n = 2)

```

* Base R approach is awkward to type and to read
* dplyr approach uses similar syntax to `filter`
* Like a SELECT in SQL

```{r results='hide'}
# base R approach to select DepTime, ArrTime, and FlightNum columns
flights[, c("DepTime", "ArrTime", "FlightNum")]

```

```{r}
# dplyr approach
select(flights, DepTime, ArrTime, FlightNum)

select(flights, DepTime, ArrTime, FlightNum) %>% dplyr::filter(DepTime == 1000)


# use colon to select multiple contiguous columns, and use `contains` to match columns by name
# note: `starts_with`, `ends_with`, and `matches` (for regular expressions) can also be used to match columns by name
select(flights, Year:DayofMonth, contains("Taxi"), contains("Delay"))

select(flights, Year:DayofMonth, contains("taxi"), contains("delay"),ends_with("ay"),matches("xiI"), starts_with("or"))


```




```{r}
iris <- dplyr::tbl_df(iris)
glimpse(iris)

dplyr::select(iris, Sepal.Width, Petal.Length, Species)
#Select columns by name or helper function.

# Regex
select(iris, contains("."))
#Select columns whose name contains a character string.

select(iris, contains("gt"))
#Select columns whose name contains a character string.

select(iris, ends_with("Length"))
#Select columns whose name ends with a character string.

select(iris, everything())
#Select every column.

select(iris, matches("dt"))
#Select columns whose name matches a regular expression.

select(iris, num_range("x", 1:5))
#Select columns named x1, x2, x3, x4, x5.


select(iris, one_of(c("Species", "Petal.Length")))
#Select columns whose names are in a group of names.

select(iris, starts_with("Sepal"))
#Select columns whose name starts with a character string.

select(iris, Sepal.Length:Petal.Width)
#Select all columns between Sepal.Length and Petal.Width (inclusive).

select(iris, -Species)
#Select all columns except Species.
```





## "Chaining" or "Pipelining"

* Usual way to perform multiple operations in one line is by nesting
* Can write commands in a natural order by using the `%>%` infix operator (which can be pronounced as "then")

```{r results='hide'}
# nesting method to select UniqueCarrier and DepDelay columns and filter for delays over 60 minutes
filter(select(flights, UniqueCarrier, DepDelay), DepDelay > 60)
```

```{r}
# chaining method
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    filter(DepDelay > 60)
```

* Chaining increases readability significantly when there are many commands
* Operator is automatically imported from the [magrittr](https://github.com/smbache/magrittr) package
* Can be used to replace nesting in R commands outside of dplyr

```{r results='hide'}
# create two vectors and calculate Euclidian distance between them
x1 <- 1:5; x2 <- 2:6
sqrt(sum((x1-x2)^2))
```

```{r}
# chaining method
(x1-x2)^2 %>% sum() %>% sqrt()
```


## arrange: Reorder rows

```{r results='hide'}
# base R approach to select UniqueCarrier and DepDelay columns and sort by DepDelay
flights[order(flights$DepDelay), c("UniqueCarrier", "DepDelay")]
```

```{r}
# dplyr approach
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(DepDelay)
```

```{r results='hide'}
# use `desc` for descending
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay))
```




## mutate: Add new variables

* Create new variables that are functions of existing variables

```{r results='hide'}
dplyr::glimpse(flights)
# base R approach to create a new variable Speed (in mph)
flights$Speed <- flights$Distance / flights$AirTime*60
flights[, c("Distance", "AirTime", "Speed")]
```

```{r}
# dplyr approach (prints the new variable but does not store it)
flights %>%
    select(Distance, AirTime) %>%
    mutate(Speed = Distance/AirTime*60)

# store the new variable
flights <- flights %>% mutate(Speed1 = Distance/AirTime*60)
```



## summarise: Reduce variables to values

* Primarily useful with data that has been grouped by one or more variables
* `group_by` creates the groups that will be operated on
* `summarise` uses the provided aggregation function to summarise each group

```{r results='hide'}
# base R approaches to calculate the average arrival delay to each destination
dplyr::glimpse(flights)
head(with(flights, tapply(ArrDelay, Dest, mean, na.rm=TRUE)))
head(aggregate(ArrDelay ~ Dest, flights, mean))
```

```{r}
# dplyr approach: create a table grouped by Dest, and then summarise each group by taking the mean of ArrDelay
flights %>%
    group_by(Dest) %>%
    summarise(avg_delay = mean(ArrDelay,  na.rm=TRUE))


# `summarise_each` allows you to apply the same summary function to multiple columns at once

flights %>%
    group_by(Dest) %>%
    summarise_each(funs(mean(., na.rm = TRUE)), ArrDelay, DepDelay, AirTime)
```

* `summarise_each` allows you to apply the same summary function to multiple columns at once
* Note: `mutate_each` is also available

```{r}
# for each carrier, calculate the percentage of flights cancelled or diverted
flights %>%
    group_by(UniqueCarrier) %>%
    summarise_each(funs(mean), Cancelled, Diverted, ArrDelay)

# for each carrier, calculate the minimum and maximum arrival and departure delays
flights %>%
    group_by(UniqueCarrier) %>%
    summarise_each(funs(min(., na.rm=TRUE), max(., na.rm=TRUE)), matches("Delay"),AirTime)
```

* Helper function `n()` counts the number of rows in a group
* Helper function `n_distinct(vector)` counts the number of unique items in that vector

```{r}
# for each day of the year, count the total number of flights and sort in descending order
dplyr::glimpse(flights)
flights %>%
    group_by(Month, DayofMonth) %>%
    summarise(flight_count = n()) %>%
    arrange(desc(flight_count))

# rewrite more simply with the `tally` function
flights %>%
    group_by(Month, DayofMonth) %>%
    tally(sort = TRUE)

# for each destination, count the total number of flights and the number of distinct planes that flew there
flights %>%
    group_by(Dest) %>%
    summarise(flight_count = n(), plane_count = n_distinct(TailNum), flighHr = sum(ActualElapsedTime, na.rm = TRUE))
```

* Grouping can sometimes be useful without summarising

```{r}
# for each destination, show the number of cancelled and not cancelled flights
flights %>%
    group_by(Dest) %>%
    select(Cancelled) %>%
    table() %>%
    head()
```


## Window Functions

* Aggregation function (like `mean`) takes n inputs and returns 1 value
* [Window function](http://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html) takes n inputs and returns n values
    * Includes ranking and ordering functions (like `min_rank`), offset functions (`lead` and `lag`), and cumulative aggregates (like `cummean`).

```{r results='hide'}
# for each carrier, calculate which two days of the year they had their longest departure delays
# note: smallest (not largest) value is ranked as 1, so you have to use `desc` to rank by largest value
head(flights, n = 2)

flights %>%
    group_by(UniqueCarrier) %>%
    select(Month, DayofMonth, DepDelay) %>%
    filter(min_rank(desc(DepDelay)) <= 2) %>%
    arrange(UniqueCarrier, desc(DepDelay))


flights %>%
    group_by(UniqueCarrier) %>%
    select(Month, DayofMonth, DepDelay, DepTime) %>%
    filter(min_rank(desc(DepDelay)) <= 10) %>%
    arrange(UniqueCarrier,Month,desc(DepDelay))
```

```{r}
# rewrite more simply with the `top_n` function
flights %>%
    group_by(UniqueCarrier) %>%
    select(Month, DayofMonth, DepDelay) %>%
    top_n(2) %>%
    arrange(UniqueCarrier, desc(DepDelay))

# for each month, calculate the number of flights and the change from the previous month
flights %>%
    group_by(Month) %>%
    summarise(flight_count = n()) %>%
    mutate(changeVs.LastMonth = flight_count - lag(flight_count))

# rewrite more simply with the `tally` function
flights %>%
    group_by(Month) %>%
    tally() %>%
    mutate(change = n - lag(n))
```


## Other Useful Convenience Functions

```{r}
# base R approach to view the structure of an object
str(flights)

# dplyr approach: better formatting, and adapts to your screen width
glimpse(flights)
```


## Connecting to Databases

* dplyr can connect to a database as if the data was loaded into a data frame
* Use the same syntax for local data frames and databases
* Only generates SELECT statements
* Currently supports SQLite, PostgreSQL/Redshift, MySQL/MariaDB, BigQuery, MonetDB
* Example below is based upon an SQLite database containing the hflights data
    * Instructions for creating this database are in the [databases vignette](http://cran.r-project.org/web/packages/dplyr/vignettes/databases.html)

```{r}
install.packages("RSQLite")
# connect to an SQLite database containing the hflights data
my_db <- src_sqlite("my_db.sqlite3")

# connect to the "hflights" table in that database
flights_tbl <- tbl(my_db, "hflights")

# example query with our data frame
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay))

# identical query using the database
flights_tbl %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay))
```

* You can write the SQL commands yourself
* dplyr can tell you the SQL it plans to run and the query execution plan

```{r}
# send SQL commands to the database
tbl(my_db, sql("SELECT * FROM hflights LIMIT 100"))

# ask dplyr for the SQL commands
flights_tbl %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay)) %>%
    explain()
```


## Resources

* [Official dplyr reference manual and vignettes on CRAN](http://cran.r-project.org/web/packages/dplyr/index.html): vignettes are well-written and cover many aspects of dplyr
* [July 2014 webinar about dplyr (and ggvis) by Hadley Wickham](http://pages.rstudio.net/Webinar-Series-Recording-Essential-Tools-for-R.html) and related [slides/code](https://github.com/rstudio/webinars/tree/master/2014-01): mostly conceptual, with a bit of code
* [dplyr tutorial by Hadley Wickham](https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a) at the [useR! 2014 conference](http://user2014.stat.ucla.edu/): excellent, in-depth tutorial with lots of example code (Dropbox link includes slides, code files, and data files)
* [dplyr GitHub repo](https://github.com/hadley/dplyr) and [list of releases](https://github.com/hadley/dplyr/releases)


< END OF DOCUMENT >



---
title: 'Going deeper with dplyr: New features in 0.3 and 0.4'
output: html_document
---

## Introduction

In August 2014, I created a [40-minute video tutorial](https://www.youtube.com/watch?v=jWjqLW-u3hc) introducing the key functionality of the dplyr package in R, using dplyr version 0.2. Since then, there have been two significant updates to dplyr (0.3 and 0.4), introducing a ton of new features.

This document (created in March 2015) covers the most useful new features in 0.3 and 0.4, as well as other functionality that I didn't cover last time (though it is not necessarily new). My [new video tutorial](https://www.youtube.com/watch?v=2mh1PqfsXVI) walks through the code below in detail.

**If you have not watched the [previous tutorial](https://www.youtube.com/watch?v=jWjqLW-u3hc)**, I recommend you do so first since it covers some dplyr basics that will not be covered in this tutorial.


## Loading dplyr and the nycflights13 dataset

Although my last tutorial used data from the hflights package, Hadley Wickham has rewritten the [dplyr vignettes](http://cran.r-project.org/web/packages/dplyr/index.html) to use the nycflights13 package instead, and so I'm also using nycflights13 for the sake of consistency.

```{r eval=FALSE}
# remove flights data if you just finished my previous tutorial
rm(flights)
```

```{r}
# load packages
suppressMessages(library(dplyr))
library(nycflights13)

# print the flights dataset from nycflights13
flights
```


## Choosing columns: select, rename

```{r}
head(flights, n = 2) # 19 Columns

# besides just using select() to pick columns...
flights %>% select(carrier, flight)

# ...you can use the minus sign to hide columns
flights %>% select(-month, -day) # Fetching only 17 columns
```

```{r results='hide'}
# hide a range of columns
flights %>% select(-(dep_time:arr_delay))

# hide any column with a matching name
flights %>% select(-contains("time"))
```

```{r}
# pick columns using a character vector of column names
cols <- c("carrier", "flight", "tailnum")
flights %>% select(one_of(cols))
```

```{r}
# select() can be used to rename columns, though all columns not mentioned are dropped
flights %>% select(tail = tailnum)

# rename() does the same thing, except all columns not mentioned are kept
flights %>% rename(tail = tailnum)
```


## Choosing rows: filter, between, slice, sample_n, top_n, distinct

```{r}
# filter() supports the use of multiple conditions
flights %>% filter(dep_time >= 600, dep_time <= 605)
```

```{r results='hide'}
# between() is a concise alternative for determing if numeric values fall in a range
flights %>% filter(between(dep_time, 600, 605))

# side note: is.na() can also be useful when filtering
flights %>% filter(!is.na(dep_time))
```


```{r}
# slice() filters rows by position
flights %>% slice(1000:1005)

# keep the first three rows within each group
flights %>% group_by(month, day) %>% slice(1:3)

# sample three rows from each group
flights %>% group_by(month, day) %>% sample_n(3)

# keep three rows from each group with the top dep_delay
flights %>% group_by(month, day) %>% top_n(3, dep_delay)

# also sort by dep_delay within each group
flights %>% group_by(month, day) %>% top_n(3, dep_delay) %>% arrange(desc(dep_delay))
```


```{r}
# unique rows can be identified using unique() from base R
flights %>% select(origin, dest) %>% unique()
```

```{r results='hide'}
# dplyr provides an alternative that is more "efficient"
flights %>% select(origin, dest) %>% distinct()

# side note: when chaining, you don't have to include the parentheses if there are no arguments
flights %>% select(origin, dest) %>% distinct
```


## Adding new variables: mutate, transmute, add_rownames

```{r}
# mutate() creates a new variable (and keeps all existing variables)
flights %>% mutate(speed = distance/air_time*60)

# transmute() only keeps the new variables
flights %>% transmute(speed = distance/air_time*60)
```


```{r}
# example data frame with row names
mtcars %>% head()

# add_rownames() turns row names into an explicit variable
mtcars %>% add_rownames("model") %>% head()

# side note: dplyr no longer prints row names (ever) for local data frames
mtcars %>% tbl_df()
```


## Grouping and counting: summarise, tally, count, group_size, n_groups, ungroup

```{r}
# summarise() can be used to count the number of rows in each group
flights %>% group_by(month) %>% summarise(cnt = n())
```

```{r results='hide'}
# tally() and count() can do this more concisely
flights %>% group_by(month) %>% tally()
flights %>% count(month)
```

```{r}
# you can sort by the count
flights %>% group_by(month) %>% summarise(cnt = n()) %>% arrange(desc(cnt))
```

```{r results='hide'}
# tally() and count() have a sort parameter for this purpose
flights %>% group_by(month) %>% tally(sort=TRUE)
flights %>% count(month, sort=TRUE)
```

```{r}
# you can sum over a specific variable instead of simply counting rows
flights %>% group_by(month) %>% summarise(dist = sum(distance))
```

```{r results='hide'}
# tally() and count() have a wt parameter for this purpose
flights %>% group_by(month) %>% tally(wt = distance)
flights %>% count(month, wt = distance)
```

```{r}
# group_size() returns the counts as a vector
flights %>% group_by(month) %>% group_size()

# n_groups() simply reports the number of groups
flights %>% group_by(month) %>% n_groups()
```

```{r}
# group by two variables, summarise, arrange (output is possibly confusing)
flights %>% group_by(month, day) %>% summarise(cnt = n()) %>% arrange(desc(cnt)) %>% print(n = 40)

# ungroup() before arranging to arrange across all groups
flights %>% group_by(month, day) %>% summarise(cnt = n()) %>% ungroup() %>% arrange(desc(cnt))
```


## Creating data frames: data_frame

`data_frame()` is a better way than `data.frame()` for creating data frames. Benefits of `data_frame()`:

* You can use previously defined columns to compute new columns.
* It never coerces column types.
* It never munges column names.
* It never adds row names. 
* It only recycles length 1 input.
* It returns a local data frame (a tbl_df).

```{r}
# data_frame() example
data_frame(a = 1:6, b = a*2, c = 'string', 'd+e' = 1) %>% glimpse()

# data.frame() example
data.frame(a = 1:6, c = 'string', 'd+e' = 1) %>% glimpse()
```


## Joining (merging) tables: left_join, right_join, inner_join, full_join, semi_join, anti_join

```{r}
# create two simple data frames
(a <- data_frame(color = c("green","yellow","red"), num = 1:3))
(b <- data_frame(color = c("green","yellow","pink"), size = c("S","M","L")))

# only include observations found in both "a" and "b" (automatically joins on variables that appear in both tables)
inner_join(a, b)

# include observations found in either "a" or "b"
full_join(a, b)

# include all observations found in "a"
left_join(a, b)

# include all observations found in "b"
right_join(a, b)

# right_join(a, b) is identical to left_join(b, a) except for column ordering
left_join(b, a)

# filter "a" to only show observations that match "b"
semi_join(a, b)

# filter "a" to only show observations that don't match "b"
anti_join(a, b)
```


```{r}
# sometimes matching variables don't have identical names
b <- b %>% rename(col = color)

# specify that the join should occur by matching "color" in "a" with "col" in "b"
inner_join(a, b, by=c("color" = "col"))
```


## Viewing more output: print, View

```{r}
# specify that you want to see more rows
flights %>% print(n = 15)
```

```{r eval=FALSE}
# specify that you want to see ALL rows (don't run this!)
flights %>% print(n = Inf)
```

```{r}
# specify that you want to see all columns
flights %>% print(width = Inf)
```

```{r eval=FALSE}
# show up to 1000 rows and all columns
flights %>% View()

# set option to see all columns and fewer rows
options(dplyr.width = Inf, dplyr.print_min = 6)

# reset options (or just close R)
options(dplyr.width = NULL, dplyr.print_min = 10)
```


## Resources

* Release announcements for [version 0.3](http://blog.rstudio.org/2014/10/13/dplyr-0-3-2/) and [version 0.4](http://blog.rstudio.org/2015/01/09/dplyr-0-4-0/)
* [dplyr reference manual and vignettes](http://cran.r-project.org/web/packages/dplyr/)
* [Two-table vignette](http://cran.r-project.org/web/packages/dplyr/vignettes/two-table.html) covering joins and set operations
* [RStudio's Data Wrangling Cheat Sheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) for dplyr and tidyr
* [dplyr GitHub repo](https://github.com/hadley/dplyr) and [list of releases](https://github.com/hadley/dplyr/releases)


## Data School

* [Blog](http://www.dataschool.io/)
* [Email newsletter](http://www.dataschool.io/subscribe/)
* [YouTube channel](http://youtube.com/dataschool)


< END OF DOCUMENT >


